---
title: "Team Data"
author: "Xavier Genelin"
---

This file will focus on getting team related data ready for use in the R Shiny application. The following libraries will be needed:
* tidyverse
* nflverse

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(nflverse)
library(fst)
```

```{r seasons}
start_season <- 2015
end_season <- 2023
```


# Offense

This section will focus on offensive data for each team. One will be broken out into running plays and the other into passing plays. At the end, both will be combined to give one dataframe for team offensive data

```{r ids}
# load player ids with their position
ids <- load_players() %>%
  select(position, gsis_id)

ids
```

```{r}
teams <- load_teams()
teams
```

```{r participation}
# load participation stats for each play
participation <- load_participation(seasons = TRUE)

participation
```

```{r pbp}
# load in play by play data and only get regular season plays
pbp <- load_pbp(seasons = c(start_season:most_recent_season())) %>%
  filter(season_type == "REG") %>%
  # get the position of the passer
  left_join(ids, by = c("passer_player_id" = "gsis_id")) %>%
  rename(passer_position = position) %>%
  # get the position of the rusher
  left_join(ids, by = c("rusher_player_id" = "gsis_id")) %>%
  rename(rusher_position = position) %>%
  # get the position of the receiver
  left_join(ids, by = c("receiver_player_id" = "gsis_id")) %>%
  rename(receiver_position = position) %>%
  # add in additional play information
  left_join(participation, by = c("game_id" = "nflverse_game_id", "play_id" = "play_id"), suffix = c("", "_remove")) %>%
  select(-contains("_remove"))

pbp
```

## Offensive Rushing

```{r rushing}
# filter the play by play data to get just rushing plays
rushing <- pbp %>%
  # don't include 2 point conversions (that will be used later) or plays that are listed as a penalty
  filter(rush_attempt == 1 & play_type_nfl != "PAT2" & play_type_nfl != "PENALTY")

rushing
```

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Get rushes inside the redzone, 10 yard line, 5 yard line and by position
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### Weekly

```{r rushing_off_weekly}
# get season rushing data of interest for each team and season
rushing_off_weekly <- rushing %>%
  group_by(posteam, season, week) %>%
  summarise(Carries = sum(rush_attempt),
            `Rush Yds` = sum(rushing_yards),
            `Rush TDs` = sum(rush_touchdown),
            `Rush Fum` = sum(fumble),
            # rushers_used = n_distinct(rusher_player_id),
            # qb_carries = sum(ifelse(rusher_position == "QB", 1, 0)),
            # qb_rushers = n_distinct(ifelse(rusher_position == "QB", rusher_player_id, NA)),
            # rb_carries = sum(ifelse(rusher_position == "RB", 1, 0)),
            # rb_rushers = n_distinct(ifelse(rusher_position == "RB", rusher_player_id, NA)),
            # wr_carries = sum(ifelse(rusher_position == "WR", 1, 0)),
            # wr_rushers = n_distinct(ifelse(rusher_position == "WR", rusher_player_id, NA)),
            # te_carries = sum(ifelse(rusher_position == "TE", 1, 0)),
            # te_rushers = n_distinct(ifelse(rusher_position == "TE", rusher_player_id, NA)),
            # qb_scramble = sum(qb_scramble),
            `Rush 20+` = sum(ifelse(rushing_yards >= 20, 1, 0)),
            `Rush 40+` = sum(ifelse(rushing_yards >= 40, 1, 0)),
            `Rush Longest` = max(rushing_yards),
            ###### will need to double check the defenders in box NA values
            # could be a weird formation from a fake punt/other weird play/formation to give a different defensive personnel
            `Rush Stacked Box` = sum(ifelse(is.na(defenders_in_box), 0, 
                                          ifelse(defenders_in_box > 7, 1, 0))),
            .groups = "drop") %>%
  mutate(`Yds/Carry` = round(`Rush Yds`/Carries, 1))
  # mutate(yard_per_car = round(rush_yards/carries, 1),
  #        #other_carries = carries - qb_carries - rb_carries - wr_carries - te_carries,
  #        qb_designed_run = qb_carries - qb_scramble)

rushing_off_weekly
```

## Offensive Passing

```{r passing}
# filter the play by play data to get just rushing plays
passing <- pbp %>%
  # don't include 2 point conversions (that will be used later) or plays that are listed as a penalty
  filter(pass_attempt == 1 & play_type_nfl != "PAT2" & play_type_nfl != "PENALTY")

passing
```

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Get passes inside the redzone, 10 yard line, 5 yard line and by position
Number of WRs, TEs, RBs used in passing game
Total number of passes to each position group
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### Weekly

```{r passing_off_weekly}
passing_off_weekly <- passing %>%
  group_by(posteam, season, week) %>%
  summarise(`Passing Yds` = sum(passing_yards, na.rm = TRUE),
            `Pass TDs` = sum(pass_touchdown),
            pass_attempt = sum(pass_attempt),
            Ints = sum(interception),
            YAC = sum(yards_after_catch, na.rm = TRUE),
            Sacks = sum(sack),
            Completions = sum(complete_pass),
            `Pass 20+` = sum(ifelse(yards_gained >= 20, 1, 0)),
            `Pass 40+` = sum(ifelse(yards_gained >= 40, 1, 0)),
            `Pass Longest` = max(yards_gained),
            ###### will need to check that the NA pass rushers are valid
            # could be a trick play out of a weird formation or fake punt type of play for a weird defensive personnel
            `Pass Against Blitz` = sum(ifelse(is.na(number_of_pass_rushers), 0, 
                                            ifelse(number_of_pass_rushers > 4, 1, 0)))) %>%
  mutate(`Pass Attempts` = pass_attempt - Sacks,
         `Completion%` = round(Completions/`Pass Attempts`, 3),
         `Yds/Att` = round(`Passing Yds`/`Pass Attempts`, 1)) %>%
  select(-pass_attempt)

passing_off_weekly 
```

```{r offense_merge}
offense_weekly <- merge(rushing_off_weekly, passing_off_weekly, by = c("posteam", "season", "week")) %>%
  rename("team" = "posteam")
```

## Defense

This section will get a lot of the same information as above, but now for the defense

## Defense Rushing

### Weekly

```{r rushing_def_weekly}
rushing_def_weekly <- rushing %>%
  group_by(defteam, season, week) %>%
  summarise(Carries = sum(rush_attempt),
            `Rush Yds` = sum(rushing_yards),
            `Rush TDs` = sum(rush_touchdown),
            `Rush Fum` = sum(fumble),
            #qb_scramble = sum(qb_scramble),
            `Rush 20+` = sum(ifelse(rushing_yards >= 20, 1, 0)),
            `Rush 40+` = sum(ifelse(rushing_yards >= 40, 1, 0)),
            `Rush Longest` = max(rushing_yards),
            ###### will need to double check the defenders in box NA values
            # could be a weird formation from a fake punt/other weird play/formation to give a different defensive personnel
            `Rush Stacked Box` = sum(ifelse(is.na(defenders_in_box), 0, 
                                          ifelse(defenders_in_box > 7, 1, 0))),
            .groups = "drop") %>%
  mutate(`Yds/Carry` = round(`Rush Yds`/Carries, 1))

rushing_def_weekly
```

## Defense Passing

### Weekly

```{r passing_def_weekly}
passing_def_weekly <- passing %>%
  group_by(defteam, season, week) %>%
  summarise(`Passing Yds` = sum(ifelse(!is.na(passing_yards), passing_yards, 0)),
            `Pass TDs` = sum(pass_touchdown),
            pass_attempt = sum(pass_attempt),
            Ints = sum(interception),
            YAC = sum(ifelse(!is.na(yards_after_catch), yards_after_catch, 0)),
            Sacks = sum(sack),
            Completions = sum(complete_pass),
            `Pass 20+` = sum(ifelse(yards_gained >= 20, 1, 0)),
            `Pass 40+` = sum(ifelse(yards_gained >= 40, 1, 0)),
            `Pass Longest` = max(yards_gained),
            ###### will need to check that the NA pass rushers are valid
            # could be a trick play out of a weird formation or fake punt type of play for a weird defensive personnel
            `Pass Against Blitz` = sum(ifelse(is.na(number_of_pass_rushers), 0, 
                                            ifelse(number_of_pass_rushers > 4, 1, 0)))) %>%
  mutate(`Pass Attempts` = pass_attempt - Sacks,
         `Completion%` = round(Completions/`Pass Attempts`, 3),
         `Yds/Att` = round(`Passing Yds`/`Pass Attempts`, 1)) %>%
  select(-pass_attempt)

passing_def_weekly
```

```{r defense_merge}
defense_weekly <- merge(rushing_def_weekly, passing_def_weekly, by = c("defteam", "season", "week")) %>%
  rename("team" = "defteam")
```

```{r}
offense_weekly <- teams %>%
  select(team_abbr, team_name) %>%
  left_join(offense_weekly, by = c("team_abbr" = "team"))

defense_weekly <- teams %>%
  select(team_abbr, team_name) %>%
  left_join(defense_weekly, by = c("team_abbr" = "team"))
```

```{r}
offense_weekly <- offense_weekly %>%
  rename_with(~ paste(., "off", sep = "_"), .cols = -(1:4))

defense_weekly <- defense_weekly %>%
  rename_with(~ paste(., "def", sep = "_"), .cols = -(1:4))

team_weekly <- offense_weekly %>%
  left_join(defense_weekly, by = c("team_abbr", "team_name", "season", "week"))
```

```{r}
team_weekly
```

```{r}
colnames(team_weekly)
```

```{r}
i <- -3
for(name in colnames(team_weekly)){
  i <- i + 1
  print(paste(i, name))
}
```


```{r}
team_season <- team_weekly %>%
  group_by(team_abbr, team_name, season) %>%
  summarise(across(c(2:5, 9, 11:16, 20:21, 24:27, 33:38, 43), sum),
            across(c(6:8, 17:19, 28:30, 39:41), max)) %>%
  ungroup() %>%
  mutate(`Yds/Carr_off` = round(`Rush Yds_off` / Carries_off, 1),
         `Yds/Carr_def` = round(`Rush Yds_def` / Carries_def, 1),
         `Yds/Att_off` = round(`Passing Yds_off` / `Pass Attempts_off`, 1),
         `Yds/Att_def` = round(`Passing Yds_def` / `Pass Attempts_def`, 1),
         `Completion%_off` = round(Completions_off / `Pass Attempts_off`, 4),
         `Completion%_def` = round(Completions_def / `Pass Attempts_def`, 4))
```

# Kicking Data
```{r kicking_weekly}
kicking_weekly <- load_player_stats(seasons = start_season:most_recent_season(), stat_type = "kicking") %>%
  filter(season_type == "REG")
```

```{r}
kicking_team_weekly <- kicking_weekly %>%
  group_by(season, team, week) %>%
  summarise(`FG Made` = sum(fg_made, na.rm = TRUE),
            `FG Att` = sum(fg_att, na.rm = TRUE),
            `FG Missed` = sum(fg_missed, na.rm = TRUE),
            `FG Blocked` = sum(fg_blocked, na.rm = TRUE),
            `FG Long` = max(fg_long),
            `FG Made 0-19` = sum(fg_made_0_19, na.rm = TRUE),
            `FG Made 20-29` = sum(fg_made_20_29, na.rm = TRUE),
            `FG Made 30-39` = sum(fg_made_30_39, na.rm = TRUE),
            `FG Made 40-49` = sum(fg_made_40_49, na.rm = TRUE),
            `FG Made 50-59` = sum(fg_made_50_59, na.rm = TRUE),
            `FG Made 60+` = sum(fg_made_60_, na.rm = TRUE),
            `FG Missed 0-19` = sum(fg_missed_0_19, na.rm = TRUE),
            `FG Missed 20-29` = sum(fg_missed_20_29, na.rm = TRUE),
            `FG Missed 30-39` = sum(fg_missed_30_39, na.rm = TRUE),
            `FG Missed 40-49` = sum(fg_missed_40_49, na.rm = TRUE),
            `FG Missed 50-59` = sum(fg_missed_50_59, na.rm = TRUE),
            `FG Missed 60+` = sum(fg_missed_60_, na.rm = TRUE),
            `PAT Made` = sum(pat_made, na.rm = TRUE),
            `PAT Att` = sum(pat_att, na.rm = TRUE),
            `PAT Missed` = sum(pat_missed, na.rm = TRUE),
            `PAT Blocked` = sum(pat_blocked, na.rm = TRUE))  %>%
  mutate(`FG%` = round(`FG Made` / `FG Att`, 4),
         `PAT%` = round(`PAT Made` / `PAT Att`, 4))

kicking_team_weekly
```


```{r kicking_season}
kicking_season <- kicking_weekly %>%
  group_by(season, team) %>%
  summarise(`FG Made` = sum(fg_made, na.rm = TRUE),
            `FG Att` = sum(fg_att, na.rm = TRUE),
            `FG Missed` = sum(fg_missed, na.rm = TRUE),
            `FG Blocked` = sum(fg_blocked, na.rm = TRUE),
            `FG Long` = max(fg_long, na.rm = TRUE),
            `FG Made 0-19` = sum(fg_made_0_19, na.rm = TRUE),
            `FG Made 20-29` = sum(fg_made_20_29, na.rm = TRUE),
            `FG Made 30-39` = sum(fg_made_30_39, na.rm = TRUE),
            `FG Made 40-49` = sum(fg_made_40_49, na.rm = TRUE),
            `FG Made 50-59` = sum(fg_made_50_59, na.rm = TRUE),
            `FG Made 60+` = sum(fg_made_60_, na.rm = TRUE),
            `FG Missed 0-19` = sum(fg_missed_0_19, na.rm = TRUE),
            `FG Missed 20-29` = sum(fg_missed_20_29, na.rm = TRUE),
            `FG Missed 30-39` = sum(fg_missed_30_39, na.rm = TRUE),
            `FG Missed 40-49` = sum(fg_missed_40_49, na.rm = TRUE),
            `FG Missed 50-59` = sum(fg_missed_50_59, na.rm = TRUE),
            `FG Missed 60+` = sum(fg_missed_60_, na.rm = TRUE),
            `PAT Made` = sum(pat_made, na.rm = TRUE),
            `PAT Att` = sum(pat_att, na.rm = TRUE),
            `PAT Missed` = sum(pat_missed, na.rm = TRUE),
            `PAT Blocked` = sum(pat_blocked, na.rm = TRUE))  %>%
  mutate(`FG%` = round(`FG Made` / `FG Att`, 4),
         `PAT%` = round(`PAT Made` / `PAT Att`, 4))

kicking_season
```

# Add Kicking data
```{r add_kicking_data}
team_weekly <- team_weekly %>%
  left_join(kicking_team_weekly, by = c("season" = "season", "team_abbr" = "team", "week" = "week"))

team_season <- team_season %>%
  left_join(kicking_season, by = c("season" = "season", "team_abbr" = "team"))
```

# Add Scoring data
```{r}
schedules <- load_schedules(seasons = start_season:most_recent_season()) %>%
  filter(game_type == "REG") %>%
  select(season, week, away_team, away_score, home_team, home_score) 

away_scores <- schedules %>%
  select(-home_team) %>%
  rename(team = away_team, 
         `Points For` = away_score,
         `Points Against` = home_score)

home_scores <- schedules %>%
  select(-away_team) %>%
  rename(team = home_team, 
         `Points For` = home_score,
         `Points Against` = away_score)

scores <- rbind(away_scores, home_scores)
scores
```

```{r}
# add in the weekly points scored and given up to the team_weekly dataset
team_weekly <- team_weekly %>%
  left_join(scores, by = c("team_abbr" = "team", "season" = "season", "week" = "week"))

# aggregate the points scored and given up
scores_agg <- scores %>%
  group_by(season, team) %>%
  summarise(`Points For` = sum(`Points For`),
            `Points Against` = sum(`Points Against`))

# add in the aggregated poitns scored and given up to the team_season dataset
team_season <- team_season %>%
  left_join(scores_agg, by = c("team_abbr" = "team", "season" = "season"))
```


# Save to fst files
```{r team_fsts}
write.fst(team_weekly, "data/team_weekly.fst")
write.fst(team_season, "data/team_season.fst")
```

